---
title: "Exploratory Data Analysis on 12 STOCKS Across Different Sectors"
author: "Onyedikachukwu Okonkwo"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, warning= FALSE, message= FALSE}
knitr::opts_chunk$set(echo = FALSE)

# SECTION 1: SETUP, LOADING PACKAGES AND DATA

# Install Packages
if(!require("tidyverse"))install.packages("tidyverse", dependencies = TRUE)
if(!require("tidyquant"))install.packages("tidyquant", dependencies = TRUE)
if(!require("e1071"))install.packages("e1071", dependencies = TRUE)
if(!require("corrplot"))install.packages("corrplot", dependencies = TRUE)
if(!require("DT"))install.packages("DT")

# Load Libraries
library(tidyverse)
library(tidyquant)
library(e1071)
library(corrplot)
library(DT)

#Create required Global Variables
tickers <- c("AAPL", "MSFT", "JPM", "V", "JNJ", "PFE", "WMT", "NKE", "XOM", "VZ", "T", "NEE")

# Load data from api using tidyquant
stock_data <- tq_get(
  tickers,
  from = "2024-11-17",
  to = "2025-11-17",
  get = "stock.prices"
)

# get DTB3 data with tryCatch to make sure that if no economic data it wont crash
risk_free_daily <- tryCatch(
  {treasury_bill_rate <- tq_get(
    "DTB3",
    get = "economic.data",
    from = (Sys.Date() - 30) # Get the last 30 days
  )
  risk_free_annual <- tail(treasury_bill_rate, 1)$price
  risk_free_daily <- risk_free_annual/ 249
  
  cat("3 Month U.S Treasury Bill Risk Free Annual: ", risk_free_annual, "\n")
  cat("Risk Free Daily: ", risk_free_daily, "\n")
  
  risk_free_daily
  },
  
  error = function(e) {
    cat("API down. Using fallback 3.9 Risk Free Annual", "\n")
    risk_free_daily = 0.01566
    
  }
  
)

# SECTION 2; CLEANING AND ORGANIZING DATA

# clean stock data and organize what I need symbol, data, adjusted and roll average
data <- stock_data %>%
  select(symbol, date, adjusted) %>%
  group_by(symbol) %>%
  mutate(return = (((adjusted - lag(adjusted))/ lag(adjusted)) * 100),
         roll_avg_20 = rollmean(adjusted, k =20, fill = NA, align = "right")
         )  # Calculate daily percentage return and rolling average
  

#optimize by removing stock data
rm(stock_data)

# SECTION 3: FUNCTIONS 

# Create Necessary Functions

# Function to plot the mean_return
line_plot <- function(ticker){ 
  
  ticker_data <- data %>% filter(symbol == ticker)
  ticker_summary <- summary_stats %>% filter(symbol == ticker)
  ticker_mean <- ticker_summary$mean_return
  ticker_median <- ticker_summary$median_return
  ticker_max <- ticker_data %>% filter(return == max(return, na.rm = TRUE))
  ticker_min <- ticker_data %>% filter(return == min(return, na.rm = TRUE))
    ggplot(ticker_data, aes(x = date, y = return)) +
    geom_line() +
    geom_point(data = ticker_max, aes(x = date, y = return), color = "purple", size = 3) +
    geom_point(data = ticker_min, aes(x = date, y = return), color = "blue", size = 3) +
    geom_hline(yintercept = ticker_mean, color = "red", linetype = "dashed", linewidth = 1) +
    geom_hline(yintercept = ticker_median, color = "green", linetype = "dashed", linewidth = 1) +
    labs(title = paste(ticker, "Return"),
         subtitle = "Red = Mean, Green = Median") +
    theme_minimal()
}

# Function to Plot Rolling average against adjusted values
rolling_average_plot <- function(ticker) {
  ticker_data <- data %>% filter(symbol == ticker) 
  ggplot(ticker_data, aes(x=date)) +
    geom_line(aes(y = adjusted), color = "black") +
    geom_line(aes(y = roll_avg_20), color = "red", linetype = "dashed") +
    labs(
      title = paste(ticker, "Price with 20-Day Moving Average"),
                    subtitle = "Black = Actual Adjusted Price, Red = 20-Day Average",
                    x = "Date",
                    y = "Price ($)") +
        theme_minimal()
}

# Validate Data
 # head(data)
 # str(data)
 # summary(data)
 # data %>% 
 #   filter(is.na(return)) %>%
 #   count(symbol)  # verify only the first day of each stock has no returns

# Mean, Median and SD (Return)

# SECTION 4: ANALYSIS

# Analyzing mean returns, volatility, skewness, kurtosis, and Sharpe Ratio
summary_stats <- data %>%
  group_by(symbol) %>%
  summarize(
    mean_return = mean(return, na.rm = TRUE),
    median_return = median(return, na.rm = TRUE), #average returns
    max_return = max(return, na.rm = TRUE),
    min_return = min(return, na.rm = TRUE),
    sd_return = sd(return, na.rm = TRUE), # Volatility higher the more volatile
    skewness = skewness(return, na.rm = TRUE), # Skewness Calculation: Performs better or worse than the mean average on a daily
    kurtosis = kurtosis(return, na.rm = TRUE),  # Kurtosis Calculation: Low kurtosis, moves predictatbly, High Kurtosis sudden drastic change
    sharpe_ratio_daily = ((mean_return - risk_free_daily)/sd_return), #calculate risk v. reward
    sharpe_ratio_annual = sharpe_ratio_daily * sqrt(249)  
  ) %>%
  arrange(desc(sharpe_ratio_annual),
          desc(mean_return),
          sd_return,
          desc(kurtosis),
          desc(skewness)
          
  )

# Analzying Correlation
corr_data <- summary_stats %>%
  select(where(is.numeric)) %>%
  cor()

# Cluster stocks for more in-depth analysis
cluster_data <- summary_stats %>%
  column_to_rownames("symbol") %>%
  select(mean_return, sd_return, skewness, kurtosis,sharpe_ratio_annual) %>%
  scale()  # get the z score for these columns

# Elbow method to determine center size
wss <- sapply(1:10, function(k) {
  set.seed(123)
  kmeans(cluster_data, centers = k)$tot.withinss
})

# Plot the elbow
plot(1:10, wss, type = "b", 
     xlab = "Number of Clusters (k)", 
     ylab = "Within-Cluster Sum of Squares",
     main = "Elbow Method for Optimal k")

set.seed(123) # makes results reproducible 

# From Elbow Method set center to 6
kmeans_result <- kmeans(cluster_data, centers =6)


# Add cluster assignment to summary_stats
summary_stats$cluster <- kmeans_result$cluster
summary_stats <- summary_stats %>%
  mutate(investor_type = case_when(
    cluster == 4 ~ "Best Bangs per Buck",
    cluster == 3 ~ "Below Average Performers",
    cluster == 2 ~ "Disaster - Avoid",
    cluster == 1 ~ "High Performers",
    cluster == 5 ~ "High Risk, Decent Return",
    cluster == 6 ~ "Average-Performers"
  ))

# Create table to analyze
cluster_summary_stats <- summary_stats %>%
  group_by(investor_type) %>%
  summarize(
    avg_mean_return = mean(mean_return),
    avg_sd_return = mean(sd_return),
    avg_skewness = mean(skewness),
    avg_kurtosis = mean(kurtosis),
    avg_sharpe = mean(sharpe_ratio_annual),
    stocks = paste(symbol, collapse = ", "),
    
    count = n()
  )%>%
  arrange(desc(avg_sharpe))


# SECTION 5: DATA VISUALIZATION

# Plot returns, and Rolling average for each stock
for(ticker in unique(data$symbol)) {
  print(line_plot(ticker))
  print(rolling_average_plot(ticker))
} # use unique to keep it from duplicating multiple of same stock

# Risk Return Scatter Plot
risk_return_scatter_plot <- ggplot(summary_stats, aes(x = sd_return, y = mean_return, color = investor_type)) +
  geom_point(size = 3) +
  geom_text(aes(label = symbol), vjust = -1, hjust = -2, size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Risk-Return Scatter Plot",
       x = "Volatility (Standard Deviation %)",
       y = "Mean Daily Return %",
       color = "Investor Type") +
  theme_minimal()

# Correlation Heat Map
corrplot(corr_data,
         method = "color",
         type = "upper",   # Show only upper triangle (less redundant)
         addCoef.col = "black",# Add correlation numbers
         tl.col = "black", # Text label color
         tl.cex = 0.8,  # Text label size
         number.cex = 0.5, # Number size
         title = "Stock Metrics Correlation Matrix",
         mar = c(1, 1, 2, 1)) # Margins for title

```

## BOTTOM LINE UP FRONT (BLUF)
Stocks: AAPL, MSFT, JPM, V, JNJ, PFE, WMT, NKE, XOM, VZ, T, NEE 
The categories are:
**Best Bang per Buck:** These are low risk, high reward, stable stocks with high returns, and are predictable for long term investment.
**High Performers:** Perform Very Well, ideal also for long term investments. 
**High Risk-Reward:** Prone to drastic rises and falls, might be suitable for day traders, and short options trading. 
**Average Performers:** Not much to write home about, decent for portfolio diversity, maybe dividens options
**Below Average Performers:** Perform below the 30 day U.S Treasure Bond, not worth the risk, buy bonds instead. 
**Disaster-Avoid:** Steady losses on average, might be worth shorting as it is in steady decline. 

After performing an Exploratory Analysis on the above stocks, the discoveries and recommendations are as follows:


```{r stocks_ratings,echo = FALSE}
stock_rating <- cluster_summary_stats %>%
  select(investor_type, stocks)
DT::datatable(stock_rating, 
              caption = "Stock Ratings",
              options = list(pageLength = 12))
```

## METHODOLOGY

Retrieved stock prices from 2024-11-18 to 2025-11-18 from Yahoo Finance via tidyquant library, as well as 3 month U.S Treasury Bond. 

Organized the data and calculated the following:

**Daily mean return or the adjusted value:** Percent change from the previous day.
**Standard Deviation:** Volatility of the stock. the higher the more volatile.
**Skewness:** The Effect outliers have on the stock.
**Kurtosis:** How extreme the changes in the stock are, the higher the more drastic.
**Sharpe Ratio:** The risk v. reward of each stock as compared to the 30 Day U.S Treasury Bond


```{r cluster_data, echo=FALSE}
DT::datatable(summary_stats,
              caption = "Stock Table",
              options = list(pageLength = 12, scrollX = TRUE),
              class = "stripe cell-border row-border hover")
```

Utilizing the elbow method, K means center is set to 6, yielding 6 different categories of the stocks by which type of investor would be interested in the stock. 

## VISUALIZATION

The Scatter Plot with all the stock performance is show below: 

```{r scatter_plot, echo = FALSE}

print(risk_return_scatter_plot)

```

**INTERPRETATION** 

The best performers not only have  higher mean daily return, but also have good stability as indicated by the lower Standard deviation of the mean returns. 

The Disaster stocks to avoid have high volatility and negative returns. 

AAPL has High Volatility and High Returns, might be worth a gamble. 

Below is the **20 day rolling average** compared to the **adjusted daily value** of each stock:

## Rolling Average Plots {.tabset}
```{r, echo=FALSE, results='asis', warning=FALSE}
for(ticker in unique(data$symbol)) {
  cat("\n###", ticker, "\n")
  print(rolling_average_plot(ticker))
  cat("\n")
}
```


Below is the **Mean Return(% change)** of each stock over Time

## Mean Average Plots {.tabset}
```{r, echo=FALSE, results='asis', warning=FALSE}
for(ticker in unique(data$symbol)) {
  cat("\n###", ticker, "\n")
  print(line_plot(ticker))
  cat("\n")
}
```


Below is the **Correlation Heat Map** which displays how each measurement correlates to one another

```{r corr_heat_map, echo=FALSE, warning=FALSE}

corrplot(corr_data,
         method = "color",
         type = "upper",   # Show only upper triangle (less redundant)
         addCoef.col = "black",# Add correlation numbers
         tl.col = "black", # Text label color
         tl.cex = 0.8,  # Text label size
         number.cex = 0.5, # Number size
         title = "Stock Metrics Correlation Matrix",
         mar = c(1, 1, 2, 1)) # Margins for title

```



## CONCLUSION

Of the 12 Stocks, JNJ and JPM are the best for long term investments followed by MSFT, WMT. AAPL has high highs, and really low lows, but great rebounds for those who can afford the wait, can turn a decent profit when buying low. T, NEE, PFE can be viable for day traders due to the predictable yet stable fluctuations between prices, but the rises and falls are not as great as MSFT.  	
V, VZ, XOM perform below average, better off just buying treasury bonds, they have better returns. 

NKE is constantly falling, might be worth a short for those that can weather that storm. 

Biggest suprise was that not all stocks in the same sector move together. 
